---
title: "20231204_DGE"
output: html_document
date: "2023-12-04"
---


```{r, setup, include=FALSE}
library(AnnotationHub)
library(ensembldb)
library(AnnotationDbi) # install from CRAN
library(clusterProfiler)
library(msigdbr)
library(pathview)
library(pheatmap)
library(colorRamp2)
library(org.Hs.eg.db) #install from CRAN
library(limma) #install from CRAN
library(edgeR)
library(DESeq2)
library(janitor)
library(statmod)
library(GO.db)
library(tidyverse) 


projDir = "/run/user/1804238067/gvfs/sftp:host=clust1-headnode,user=lythgo02/mnt/nas1-data/jblab/group_folders/emily_lythgoe/20230724_EcoTyper_2nd_year"
knitr::opts_knit$set(root.dir = projDir) #sets location for output files to projDir 
knitr::opts_chunk$set(echo = TRUE) #setting default option for all code chunks 
```

# Differential Gene Expression Analysis on TCGA Ovarian Cohort
##BRCA1/2 loss vs non-BRCA1/2 HRD high vs HRD low (assumed HR proficient)

###Requires output from 20231204_datawrangling_DGE.Rmd


Because I will filter out genes with low expression before DGE, makes sense
to filter the genes before annotation.This should reduce the number of genes 
with duplicate IDs allowing me to manually check any duplicates and select 
which to keep. 

##Filter genes by expression

```{r}
rawCounts_filtered <- as.matrix(rawCounts_final[,-1]) # remove gene_name col
cpmCounts <- cpm(rawCounts_filtered)
#cpm() calculates counts per million for normalisation

keepGenesCPM <- rowMax(cpmCounts) >= 4
#logical vector of genes to keep based on expression levels

filCounts <- rawCounts_filtered[keepGenesCPM,] # 22250 genes remain
keepGenesExpr <- filterByExpr(filCounts, 
                              group = meta_rawCounts_matched$contrast_group)
genesToKeep <- names(keepGenesExpr)[keepGenesExpr]
length(genesToKeep)
# [1] 20981
```

We may still filter out more genes due to annotation issues, but for now this
is all the genes we might end up using.

##Generate an annotation

```{r geneAnnotation}
colGene <- c("ENTREZID", "SYMBOL", "GENETYPE", "ENSEMBL") 

#query org.Hs.eg.db gene annotation database using the Ensembl Gene IDs as keys, 
#retrieving  the specified columns (colGene)
geneIdMap <- AnnotationDbi::select(org.Hs.eg.db, 
                                  keys = genesToKeep,
                                  columns = colGene, 
                                  keytype = "ENSEMBL") 

sum(is.na(geneIdMap$ENTREZID))
# [1] 2787

dupEntrez <- geneIdMap %>% 
  filter(!is.na(ENTREZID)) %>%
  filter(n() > 1, .by = ENTREZID)
#filter missing and non-unique rows by ENTREZID

length(unique(dupEntrez$ENTREZID))
# [1] 33

```

33 = small enough number to manually check via Ensembl and try to decide 
which is the most appropriate. Can also check expression levels:

```{r, check expression for duplicates}

SPATA13 <- dupEntrez %>% filter(ENTREZID==221178) %>% pull(ENSEMBL)

#density plot of logCPM for each ENSEMBL ID of SPATA13
 cpmCounts[rownames(cpmCounts)%in%SPATA13,] %>%
    as.data.frame() %>%
    rownames_to_column("GeneID") %>%
    pivot_longer(names_to = "Sample", values_to = "CPM", -GeneID) %>%
    mutate(logCPM = log2(CPM + 0.1)) %>%
    ggplot(aes(x = logCPM)) +
        geom_density(aes(colour = GeneID))
```
given that these are overlapping loci (see Ensembl) I'd select the gene with 
the higher expression - ENSG00000182957 


Or for RABL2A we have 

```{r}
dupEntrez %>% filter(SYMBOL=="RABL2A")    
```
But Emsembl says that ENSG00000079974 should be RABL2B, which is ENTREZ 11158:
https://www.ncbi.nlm.nih.gov/gene/11158

In this case we have a duplicated Ensembl ID:

```{r}
filter(geneIdMap, ENTREZID==11159
)
filter(geneIdMap, ENSEMBL=="ENSG00000079974") 
```

So we need to delete the entry that is 
        ENSEMBL ENTREZID SYMBOL       GENETYPE
ENSG00000079974    11159 RABL2A protein-coding


I went through the dupEntrez list and checked each ID. Most were incorrectly 
labelled, some had multiple ID for the same locus in which case I checked which 
had the higher count value. For RNVU1-29 and RNVU1-18 I picked at random due to 
lack of info. This left me with the genes in Entreztokeep.csv
```{r}
#remove from geneIdMap
EntrezToKeep <- read_csv("./Entreztokeep.csv") %>% 
  select(-"Actual ENSEMBL match",-"Annotation",-"Density plot readout") %>% 
  as.data.frame() 
EntrezToKeep$ENTREZID <- as.character(EntrezToKeep$ENTREZID)

#create a df of the genes to remove from the larger geneIDmap  
EntrezToDiscard <- anti_join(dupEntrez,EntrezToKeep, by= c("ENSEMBL","ENTREZID"))

#remove from geneIdMap
geneIdMapFil <- anti_join(geneIdMap,EntrezToDiscard, by= c("ENSEMBL","ENTREZID"))


```


```{r}
dupEnsembl <- geneIdMapFil %>% 
  filter(!is.na(ENTREZID)) %>%
  filter(n() > 1, .by = ENSEMBL)
length(unique(dupEnsembl$ENSEMBL))
# [1] 145 after filtering out duplicated ENTREZID
```

Keep the first of each duplicated ENSEMBL ID.

```{r}
#keep the first instance of each duplicated ENSEMBL ID
geneIdMapFil <- geneIdMapFil %>% 
        distinct(ENSEMBL, .keep_all=TRUE)

  
sum(duplicated(geneIdMapFil$ENSEMBL))
# [1] 0
# There's no remaining duplicated ENSEMBL ids.

```
Dont throw out ones with missing Entrez ID as this doesn't matter for GSEA, 
just use the ENSEMBL ID and then you will have the same gene list for the GSEA 
as you do for the DGE.

##Alternate method of mapping using AnnotationHub

May be a more suitable approach as AnnotationHub only returns 18 duplicates 
(after filtering out low expression genes) and won't annotate pseudo/ncRNA etc 
if there is a protein coding version available.


```{r usingAnnotationHub, eval=FALSE}
# Create an AnnotationHub object for Homo sapiens
ah <- AnnotationHub()

# Create a query to retrieve gene annotation data from AnnotationHub
humanEnsDb <- query(ah, c("EnsDb", "Homo sapiens", "109"))[[1]]
annotations <- genes(humanEnsDb, return.type = "data.frame")

annot <- annotations %>%
   select(gene_id, gene_name, entrezid) %>%
   filter(gene_id %in% genesToKeep)

length(unique(annot$gene_id)) 

sum(is.na(annot$entrezid)) 

annot %>%
    filter(!is.na(entrezid)) %>%
    filter(n() > 1, .by = entrezid)
```

There are only a few duplicated ENTREZ ids, so this would be easy to sort out.
Also, the ones that we are missing here but get using the org.Hs.eg.db method
are not that interesting:

```{r eval=FALSE}
annot %>%
    dplyr::filter(is.na(entrezid)) %>%
    inner_join(geneIdMapFil, by = c(gene_id = "ENSEMBL")) %>%
    as_tibble() %>%
    count(GENETYPE)
```

I will continue with the annotation generated by the org.Hs.eg.db methods for
now.

```{r prepare_dge_list_object}
counts <- rawCounts_final[geneIdMapFil$ENSEMBL,] %>%
    select(-gene_name) %>%
    as.matrix()

dge <- DGEList(counts = counts, 
                 genes = geneIdMapFil, 
                 samples = meta_rawCounts_matched, 
                 group = meta_rawCounts_matched$contrast_group) 
```


# Process all contrast groups together

```{r, visualise count distribution}

#check for samples that have more counts on average than other samples 
boxplot(log(dge$counts + 0.5), 
        xaxt = "n", 
        col = "turquoise4", 
        xlab = "Samples") 
boxplot(cpm(dge$counts, log = TRUE), 
        xaxt = "n", 
        col = "turquoise4", 
        xlab = "Samples")

plotSparsity(dge$counts, normalized = TRUE) 
# normalised to account for variations in library size to avoid sequencing bias
```


**Normalisation**

```{r}
# calculate normalisation factors (does not normalise)
dgeNorm <- calcNormFactors(dge)

hist(dgeNorm$samples$norm.factors)

#or with ggplot
# Extract normalization factors from the DESeqDataSet object
norm_factors <- dgeNorm$samples$norm.factors

# Create a histogram using ggplot2
ggplot(data.frame(NormFactors = norm_factors), aes(x = NormFactors)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Normalization Factors",
       x = "Normalization Factors",
       y = "Frequency") +
  theme_minimal()
```

##Check for interactions 

## MDS plots {.tabset}
Multidimensional scaling plot of distances between gene expression profiles
This function uses MDS to produce a principal coordinate (PCoA) or principal
component (PCA) plot showing the relationships between samples. Distance 
between spots (samples) on plot represent log2 fold changes between samples. 
Use to identify interacting variables that you may want to control for in DE.



```{r getMDSdata, include = FALSE}
mdsDat <- plotMDS(dgeNorm, plot = FALSE)
#only need to run plotMDS once for all factors you want to plot (all contained within mdsDat object)

plotDat <- as_tibble(dgeNorm$samples) %>%
    mutate(X = mdsDat$x) %>%
    mutate(Y = mdsDat$y)

plotDat
```

## Contrast Group

```{r}
ggplot(plotDat, aes(x = X, y = Y)) +
        geom_point(aes(colour = contrast_group)) +
  labs(title = "Contrast Group")
```

## Stage

```{r}
ggplot(plotDat, aes(x = X, y = Y)) +
        geom_point(aes(colour = figo_stage))
```

## Ecotype

```{r}
ggplot(plotDat, aes(x = X, y = Y)) +
        geom_point(aes(colour = carcinoma_ecotype))
```
Some separation of ecotypes but not correcting for this as it is confounded
with gene expression (used to generate the CE classification) and the contrasts 
## Site

```{r}
ggplot(plotDat, aes(x = X, y = Y)) +
        geom_point(aes(colour = site_of_resection_or_biopsy))
```

**Define design matrix**

```{r}

modelMatrix <- model.matrix(~ 0 + contrast_group, data=dgeNorm$samples)
colnames(modelMatrix) <- str_remove(colnames(modelMatrix), "contrast_")
```

**Voom transformation** 

```{r}
voom <- voom(dgeNorm, modelMatrix, plot = TRUE) 
# transforms counts to log2 CPM based on the normalisation factors calculated
# earlier & fits a linear model to the log2 CPM for each gene (row-wise),
# calculates the residuals, and fits a smoothed curve to the square root of the
# residual standard deviation by average expression. The smoothed curve is used
# to obtain weights for each gene and sample, which are then passed into limma
# for DE, taking into account the variability of each gene.
```


**Fit linear model**

```{r}
# fitting a linear model using weighted least squares for each gene
fit <- lmFit(voom, modelMatrix) # now run DEA
head(coef(fit)) 
#display coefficients used for hypothesis testing of sig. up/down regulated genes 

```

**Define contrast**

BRCA1/2 loss is the baseline for each contrast

```{r}
# comparison between groups using the fitted linear model gives log-fold changes
contr <- makeContrasts(groupHRD_high - groupBRCA1_2_loss, 
                       groupHRD_low - groupBRCA1_2_loss, 
                       levels = modelMatrix) 
#  groupHRD_high - groupBRCA1_2_loss, # Coefficient 1
 # groupHRD_low - groupBRCA1_2_loss,  # Coefficient 2
contr
#-1 = reference
#0 = not included in comparison
#1 = group to compare against reference

contrastsfit <- contrasts.fit(fit, contrasts = contr) 
# fit contrasts to model & calculate coefficients 

fittedContrasts <- eBayes(contrastsfit, robust = TRUE) 

```
###eBayes() Robust emperical bayes for statistical testing to account for heteroscedasticy 
in the data. Helps to obtain more accurate p-values and effect size estimates.

**Extract to contrasts and statistical testing**
The output from topTable() includes:
-logFC: the log fold-change between cases and controls
-t: the t-statistic for DE
-P.Value: for DE (not adjusted for multiple testing)
-adj.P.Val: p-value adjusted for multiple testing. 
 Default adjustment method = Benjamini-Horchberg

```{r}
results_HRDhigh <- topTable(fittedContrasts, 
                            coef= "groupHRD_high - groupBRCA1_2_loss", 
                            sort.by = "P", n = Inf) 
head(results_HRDhigh, 20) # look at top 20 DE genes (sorted by p value), n=Inf means retreive all DE genes (no limit on number)

up_genes_BRCAloss_HRDhigh <- results_HRDhigh %>%
  dplyr::filter(adj.P.Val < 0.05, logFC >= 1.0) %>% 
  write_csv("Output/Data/20231211_up_genes_BRCAloss_HRDhigh.csv")
  
down_genes_BRCAloss_HRDhigh <- results_HRDhigh %>%
  dplyr::filter(adj.P.Val < 0.05, logFC <= -1.0) %>% 
  write_csv("Output/Data/20231211_down_genes_BRCAloss_HRDhigh.csv")
  
allDEgenes_BRCAloss_HRDhigh <- results_HRDhigh %>%
  dplyr::filter(adj.P.Val < 0.05, abs(logFC) >= 1.0) %>% 
  write_csv("Output/Data/20231211_DEgenes_BRCAloss_HRDhigh.csv")

results_HRDlow <- topTable(fittedContrasts, coef= "groupHRD_low - groupBRCA1_2_loss", sort.by = "P", n = Inf)
head(results_HRDlow, 20)

up_genes_BRCAloss_HRDlow <- results_HRDlow %>%
  dplyr::filter(adj.P.Val < 0.05, logFC >= 1.0) %>% 
   write_csv("Output/Data/20231211_up_genes_BRCAloss_HRDlow.csv") 

down_genes_BRCAloss_HRDlow <- results_HRDlow %>%
  dplyr::filter(adj.P.Val < 0.05, logFC <= -1.0) %>% 
  write_csv("Output/Data/20231211_down_genes_BRCAloss_HRDlow.csv")
  
allDEgenes_BRCAloss_HRDlow <- results_HRDlow %>%
  dplyr::filter(adj.P.Val < 0.05, abs(logFC) >= 1.0) %>% 
  write_csv("Output/Data/20231211_DEgenes_BRCAloss_HRDlow.csv")
```

**Venn Diagram**
```{r}
  venn_list <- list(
      up_in_HRD = as.character(up_genes_BRCAloss_HRDhigh$SYMBOL),
      up_in_HRP = as.character(up_genes_BRCAloss_HRDlow$SYMBOL),
      down_in_HRD = as.character(down_genes_BRCAloss_HRDhigh$SYMBOL),
      down_in_HRP = as.character(down_genes_BRCAloss_HRDlow$SYMBOL))
  
 
vennDat <- ggvenn::ggvenn(venn_list, set_name_size = 3)   

vennDat + labs(title = "In comparison to BRCA1/2 loss")

```
As we would expect, lots more genes up/down regulated in the HR proficient group 
in comparison to the BRCA1/2-mediated HRD group than we see when we compare the 
non BRCA1/2-mediated HRD (suspected) group to the BRCA1/2-HRD group.

**Sanity checks**
```{r}
hist(results_HRDhigh$P.Value)

hist(results_HRDlow$P.Value)
```

**Just looking at BRCA1/2 vs HRD low**
```{r}

# Total number of DE genes based on adjusted p value < 0.05
length(which(results_HRDlow$adj.P.Val < 0.05))

# Total number of DE genes based on adjusted p value < 0.05 and |logFC| (absolute log fold-change) >= 1.0
length(which(results_HRDlow$adj.P.Val < 0.05 & abs(results_HRDlow$logFC) >= 1.0 ))

# Total number of DE genes based on adjusted p value < 0.01 and |logFC| >= 1.0
length(which(results_HRDlow$adj.P.Val < 0.01 & abs(results_HRDlow$logFC) >= 1.0 ))

dtTab <- decideTests(fittedContrasts, p.value = 0.05, lfc = 1.0) 
#makes DE gene calls based on a specified p-value and log-fold change threshold 
summary(dtTab)
```


**Plots for DE contrast: BRCA1/2 loss vs HRD low**

```{r}
#to plot a measure of significance against fold change 
volcanoplot(
  fittedContrasts,             
  coef = "groupHRD_low - groupBRCA1_2_loss", # Specify which contrast to visualize
  highlight = 10,               # The number of top genes to highlight 
  names = fittedContrasts$genes$SYMBOL,    
  main = "Volcano Plot"        # Plot title 
)

#x axis - upregulation (positive values) and downregulation (negative) of genes 
#y-axis - statistical significance 
#each gene = a dot

```

```{r MA plot with ggplot}
#MA plot displays log-fold change against the average log expression for each gene 
#build necessary data into dataframe for use with ggplot
#require adjusted p-value/FDR
#greppel = handy package to push labels around geom_text_repel
library(ggrepel)

MAplotlow <- results_HRDlow %>%  arrange(desc(adj.P.Val))
#arrange so we can label the top 10 
MAplot_HRDlow <- ggplot(MAplotlow, aes(x = AveExpr, y = logFC)) +
    geom_point(aes(colour = adj.P.Val < 0.05), size = 1) +
    geom_text_repel(data = ~top_n(.x, 10, wt = -adj.P.Val), #select top 10 sig genes for labeling by -ve adj.p.val
                    aes(label = SYMBOL),
                    size=3.3,
                    min.segment.length = unit(0, 'lines'),
                    box.padding = 0.5) +
    labs(
        x = "Average Log-Expression",
        y = "Log-Fold Change",
        title = "MA Plot: HR Proficient vs BRCA1/2 Loss",
        color = "FDR < 0.05"
    ) + 
   theme_classic()


MAplothigh<- results_HRDhigh %>% arrange(desc(adj.P.Val))
MAplot_HRDhigh <- ggplot(MAplothigh, aes(x=AveExpr, y=logFC)) +
  geom_point(aes(colour = adj.P.Val < 0.05), size=1) + 
               geom_text_repel(data= ~top_n(.x, 10, wt=-adj.P.Val), 
                               aes(label=SYMBOL),
                               size=3.3,
                                min.segment.length = unit(0, 'lines'),
                                box.padding = 0.5) +
    labs(
        x = "Average Log-Expression",
        y = "Log-Fold Change",
        title = "MA Plot: Non-BRCA1/2 HRD vs BRCA1/2 Loss",
        color = "FDR < 0.05"
    ) +
  theme_classic()

MAplot_HRDlow
MAplot_HRDhigh
```
ENSG00000280303 = E2F1-regulated inhibitor of cell death lncRNA, missing 
annotation so filtered out from heatmap along with 27 others for the 
HRDlow comparison. All those missing annotation appear to be non-coding 
or novel genes.


```{r, pheatmap - HRDlow vs BRCA1/2loss}

# Select the top genes based on adj.P.Val
  topGenesHRDlow <- results_HRDlow %>% 
  top_n(500, wt= -adj.P.Val) %>% 
  pull(ENSEMBL)

plotDatlow <- voom$E[topGenesHRDlow,] 

#column for annotation
col_annot1 <- data.frame(Status=voom$targets$group,
                    Ecotype=voom$targets$carcinoma_ecotype,
                    submitter_id=voom$targets$submitter_id) %>% 
  filter(Status != "HRD_high") %>% 
  mutate(submitter_id = str_replace_all( submitter_id,"-", "."))

rownames(col_annot1) <- col_annot1$submitter_id 
#leave column names in additional column for clustering later

col_annot1 <- col_annot1 %>% mutate(across(Ecotype, ~replace_na(.x, "NA")))


#data to plot, filtering out the HRD high cases
plotDatlow <- data.frame(plotDatlow) %>% 
    select(any_of(col_annot1$submitter_id)) %>% 
    as.matrix()
#z-scale the data (results = number of SD the GE is from the mean for that gene 
#across all samples)
  z.mat1 <- t(scale(t(plotDatlow), center=TRUE, scale=TRUE))
  
  
#cluster the genes with distance matrix (allows for separation )   
  hc1 <- hclust(dist(z.mat1)) 
#tried plotting the heatmap without clustering first and saw clear separation 
#into two groups so just pulling out those groups by clustering first and then 
#adding the resulting groups as annotation  

  cuthc1 <- cutree(hc1, k=2) 
#cut tree into 2 clusters, can specify height to cut tree or number of clusters
 #1's and 2's indicate the cluster that the genes belong to

  #row annotation  
gene_clusterlow <- data.frame(Cluster = cuthc1)

#colours for annotation
colour_list1 <- list(Cluster = c("1" = "cadetblue1", "2" = "darkcyan"),
                    Status = c("BRCA1_2_loss" = "pink", 
                              "HRD_low" = "purple"),
                             #  "HRD_high" = "cornflowerblue"),
                    Ecotype= c("CE1"="darkorange", "CE2"="yellow", 
                               "CE3"="darkolivegreen1", "CE4"="deeppink3", 
                               "CE5"="pink", "CE6"="darkslateblue", 
                               "CE7"="red", "CE8"="darkviolet", 
                               "CE9"="darkseagreen3", "CE10"="#00BFC4" , 
                               "NA"="grey"))

maplow<- pheatmap(z.mat1, 
         cluster_rows = TRUE,  # Hierarchical clustering for rows
         cluster_cols = TRUE, # No clustering for columns
         color = colorRampPalette(c("blue", "white", "red"))(100), 
         main = "Top 500 DE Genes HRD low vs BRCA1/2 loss", # Heatmap title
         annotation_row = gene_clusterlow,
         annotation_col = col_annot1[,1:2],
         annotation_colors = colour_list1,
         show_rownames = FALSE,  
         show_colnames = FALSE, 
         cutree_rows = 2, 
         cutree_cols = 3,
         annotation_legend = TRUE,
         annotation_names_col = FALSE,
         annotation_names_row = FALSE)

IDclusterslow <- data.frame(
                       cluster = cutree(maplow$tree_col, k=3)) %>% 
  rownames_to_column() %>% 
  mutate(rowname = str_replace_all(rowname,"\\.", "-")) %>% 
  rename(submitter_id = rowname) %>% 
 write_csv("./Output/Data/20240117_HRDlowvsBRCA12loss_IDclusters.csv")
```
Plot all patients on one plot to see if you still get the small cluster (which you do)

```{r, pheatmap - HRDhigh vs BRCA1/2loss}

# Select the top genes based on adj.P.Val
  topGenesHRDhigh <- results_HRDhigh %>% 
  top_n(500, wt= -adj.P.Val) %>% 
  pull(ENSEMBL)

plotDathigh <- voom$E[topGenesHRDhigh,]

#column annotation
col_annot2 <- data.frame(Status=voom$targets$group,
                    Ecotype=voom$targets$carcinoma_ecotype,
                    submitter_id=voom$targets$submitter_id) %>% 
  filter(Status != "HRD_low") %>% 
  mutate(submitter_id = str_replace_all( submitter_id,"-", ".")) 

rownames(col_annot2) <- col_annot2$submitter_id

#data to plot, filtering out the HRDlow cases
plotDathigh <- data.frame(plotDathigh) %>% 
    select(any_of(col_annot2$submitter_id)) %>% 
    as.matrix()
#z-scale the data (results = number of SD the GE is from the mean for that gene 
#across all samples)
  z.mat2 <- t(scale(t(plotDathigh), center=TRUE, scale=TRUE))
  
  
#cluster the genes with distance matrix (allows for separation )   
  hc2 <- hclust(dist(z.mat2)) 
#tried plotting the heatmap without clustering first and saw clear separation 
#into two groups so just pulling out those groups by clustering first and then 
#adding the resulting groups as annotation  

  cuthc2 <- cutree(hc2, k=2) 
#cut tree into 2 clusters, can specify height to cut tree or number of clusters
 #1's and 2's indicate the cluster that the genes belong to
  
gene_clusterhigh <- data.frame(Cluster = cuthc2)
colour_list2 <- list(Cluster = c("1" = "cadetblue1", "2" = "darkcyan"),
                    Status = c("BRCA1_2_loss" = "pink",
                              # "HRD_low" = "purple",
                              "HRD_high" = "cornflowerblue"),
                    Ecotype= c("CE1"="darkorange", "CE2"="yellow", 
                               "CE3"="darkolivegreen1", "CE4"="deeppink3", 
                               "CE5"="pink", "CE6"="darkslateblue", 
                               "CE7"="red", "CE8"="darkviolet", 
                               "CE9"="darkseagreen3", "CE10"="#00BFC4" , 
                               "NA"="grey"))
maphigh <- pheatmap(z.mat2, 
         cluster_rows = TRUE,  # Hierarchical clustering for rows
         cluster_cols = TRUE, # No clustering for columns
         color = colorRampPalette(c("blue", "white", "red"))(100), 
         main = "Top 500 DE genes HRD high vs BRCA1/2 loss", # Heatmap title
         annotation_row = gene_clusterhigh,
         annotation_col = col_annot2[,1:2],
         annotation_colors = colour_list2,
         show_rownames = FALSE,  
         show_colnames = FALSE, 
         cutree_rows = 2, 
         cutree_cols = 3,
         annotation_legend = TRUE,
         annotation_names_col = FALSE,
         annotation_names_row = FALSE,
         legend_breaks = c(-2, 0, 2))
        # filename="HRDhigh_vs_BRCA1/2.pdf")
         #legend=FALSE)

IDclustershigh <- data.frame(
                       cluster = cutree(maphigh$tree_col, k=3)) %>% 
  rownames_to_column() %>% 
  mutate(rowname = str_replace_all(rowname,"\\.", "-")) %>% 
  rename(submitter_id = rowname) %>% 
 write_csv("./Output/Data/20240117_HRDhighvsBRCA12loss_IDclusters.csv")
```
##Check for biological functions that are over-represented in the gene clusters

Over-Representation Analysis (ORA): With listed DE genes, identify the gene sets 
in which the DE genes are over or under represented. The significance of each 
pathway is measured by calculating the probability that the observed number of 
DE genes in a given pathway were simply observed by chance.


###Clusterprofiler (Yu et al. 2012) Supports direct online access of 
current KEGG database (KEGG: Kyoto Encyclopedia of Genes and Genomes)

##KEGG on DE genes with no clustering: 

```{r, KEGG enrichment analysis - sig DE genes (not heatmap results)}

search_kegg_organism('human', by='common_name') #kegg code = hsa

sigGenes_HRDhigh <- allDEgenes_BRCAloss_HRDhigh %>% 
   pull(ENTREZID) %>% 
  as.numeric()

sigGenes_HRDlow <- allDEgenes_BRCAloss_HRDlow %>% 
   pull(ENTREZID) %>% 
  as.numeric()

 keggReshigh <- enrichKEGG(gene= sigGenes_HRDhigh, organism = 'hsa')
 
 keggReslow <- enrichKEGG(gene= sigGenes_HRDlow, organism = 'hsa')
 
 as_tibble(keggReslow)
 browseKEGG(keggReslow, 'hsa04657') # to visualise one of the hits in browser
 
``` 
When we take the list of sig DE genes for each comparison and perform ORA:
No pathways for HRD high vs BRCA1/2 loss
3 pathways for HRD low vs BRCA1/2 loss including IL17

###https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9827088/ IL-17 (from T-helpers?) 
upregulates MTA1 mRNA and protein expression to promote ovarian cancer 
migration and invasion

Try an increase the number of genes we can look at by taking the results from 
the heatmaps where I took the top 500 DE genes for each comparison and clustered. 
2 gene clusters, perform ORA on each cluster:


```{r, enrichment analysis/ORA, HRD low vs BRCA1/2 loss}
#take the pheatmap gene clusters that are DE between HRDlow vs BRCA1/2 loss 
gene_clusterlow <- gene_clusterlow %>% rownames_to_column("ENSEMBL")

#separate out 1st gene cluster
gene_clusterlow_ORA1 <- 
  left_join(gene_clusterlow, results_HRDlow) %>% 
#  filter(cluster==1) %>% 
  pull(ENTREZID) %>% 
  as.numeric()

#separate out second gene cluster
gene_clusterlow_ORA2 <- 
  left_join(gene_clusterlow, results_HRDlow) %>% 
  filter(Cluster==2) %>% 
  pull(ENTREZID) %>% 
  as.numeric()

#KEGG enrichment analysis

search_kegg_organism('human', by='common_name') #kegg code = hsa

#1st gene cluster
 keggReslow1 <- enrichKEGG(gene= gene_clusterlow_ORA1, organism = 'hsa')
#2nd gene cluster 
  keggReslow2 <- enrichKEGG(gene= gene_clusterlow_ORA2, organism = 'hsa')
 
 as_tibble(keggReslow1)
 as_tibble(keggReslow2)
 browseKEGG(keggReslow1, "hsa00190")
``` 
For HRD low vs BRCA1/2 loss:
Gene cluster 1 gives a list of 5 pathways
Gene cluster 2 gives a list of 7 pathways

Oxidative phosphorylation = top hit for each i.e Metabolic differences between 
BRCA1/2 loss vs HRD low  Studies in breast and ovarian cancer HRD models depict 
a metabolic shift that includes enhanced expression of the OXPHOS pathway and 
its key components and a decline in the glycolytic Warburg phenotype. Hence, 
HRD cells are more sensitive to metformin and NAD+ concentration changes.

Paper to read: https://translational-medicine.biomedcentral.com/articles/10.1186/s12967-023-04349-3

```{r, enrichment analysis/ORA, HRD high vs BRCA1/2 loss - comparing gene clusters from heatmap}
#take the pheatmap gene clusters that are DE between HRDlow vs BRCA1/2 loss 
gene_clusterhigh <- gene_clusterhigh %>% rownames_to_column("ENSEMBL")

#separate out 1st gene cluster
gene_clusterhigh_ORA1 <- 
  left_join(gene_clusterhigh, results_HRDhigh) %>% 
#  filter(cluster==1) %>% 
  pull(ENTREZID) %>% 
  as.numeric()

#separate out second gene cluster
gene_clusterhigh_ORA2 <- 
  left_join(gene_clusterhigh, results_HRDhigh) %>% 
  filter(Cluster==2) %>% 
  pull(ENTREZID) %>% 
  as.numeric()

#KEGG enrichment analysis

search_kegg_organism('human', by='common_name') #kegg code = hsa

#1st gene cluster
 keggReshigh1 <- enrichKEGG(gene= gene_clusterhigh_ORA1, organism = 'hsa')
#2nd gene cluster 
  keggReshigh2 <- enrichKEGG(gene= gene_clusterhigh_ORA2, organism = 'hsa')
 
 as_tibble(keggReshigh1)
 as_tibble(keggReshigh2)
``` 


No pathways returned for HRD high vs BRCA12 loss

I took the patient clusters from each heatmap and fed them into cBioportal to look at survival and clinical differences. No significant differences in survival outcomes. Aneuploidy states differed significantly between the clusters for both sets (the HRDhigh-BRCA12loss and HRDlow-BRCA12loss). Plots and stats in cBioportal folder in Outputs.
```{r, compare ecotypers in patient clusters from heatmaps, HRDlow vs BRCA1/2 loss}

IDclusterslow <- IDclusterslow %>% left_join(meta_rawCounts_matched)

ecotypes <- IDclusterslow %>%
  tabyl(cluster,  carcinoma_ecotype) %>% 
as.data.frame() %>% 
  select(cluster, CE1, CE2, CE3, CE4, CE5, CE6, CE7, CE8, CE9, CE10,-"NA_")

# Calculate proportions within each cluster for each CE
ecotypes_proportions <- ecotypes %>%
  mutate_at(vars(starts_with("CE")), function(x) x / rowSums(select(., starts_with("CE"))))


 ecotypes_long <- tidyr::gather(ecotypes_proportions, key = "CE", value = "proportion", -cluster)
 ecotypes_long$CE <- factor(ecotypes_long$CE, levels = c("CE1", "CE2","CE3","CE4","CE5","CE6","CE7","CE8","CE9","CE10"))
 
# Create a stacked bar chart
ggplot(ecotypes_long, aes(x = cluster, y = proportion, fill = CE)) +
     geom_bar(stat = "identity", position = "stack") +
     labs(title = "Counts of ecotypes per patient cluster",
          x = "Cluster",
          y = "Count") +
     scale_fill_brewer(palette = "Paired") +
     theme_minimal()
```

To do: 
Aneuploidy differences point towards CIN, assign the dominant signature from the pancancer paper as labels and look for differences between the clusters.
Compare the ecotype proportions between the clusters (the smallest clusters seem to be enriched for CE8/7)
Could look at the type of BRCA1/2 alteration
Use tools to look at immune infiltration in gene expression data (although this is essentially what ecotyper does so jusst look into this more)

